Towards Explainability of Machine Learning Models in Insurance Pricing

### 2019-10-25 manuscript/manuscript.Rmd
copyedits

Proposed by @hiltonConversationalProcesses1990 and interpreted by @millerExplanationArtificial2017 in the context of ML, model explanations are conversations or social interactions. One consequence of this identification is that explanations need to be *relevant* to the *audience*. In developing, filing, and operationalizing a pricing model, one needs to accommodate a variety of stakeholders, each of whom has a different set of questions, assumptions, and technical capacity. First, there are internal stakeholders at the company, which includes management and underwriters. While some of the individuals in this audience may be technical, they are likely less familiar with predictive modeling techniques than the actuaries and data scientists who build the models. Next, we have the regulators, who may have limited resources to review the models, and will focus on a specific list of questions motivated by statute. Finally, we have the public, who has an interest (perhaps more so than the other parties) because their lives are directly affected by the premiums charged.

It is interesting to note that, the modelers, who are most familiar with the models, tend to be same people designing and communicating the explanations. This poses a challenge that @millerExplainableAI2017 calls "inmates running the asylum", where the modelers design explanations for *themselves* rather than the intended audience. For example, they may be interested in technical questions like extrapolation behavior and shape the explanations accordingly, which may be lost on the prospective policyholder.

@@

Now that we have established a framework for model interpretation in the form of asking and answering relevant questions, we demonstrate examples of such exchanges via an illustrative case study. Analytically, our starting point is a fitted deep neural network model for predicting loss costs. As the modeling details are of secondary importance, they are available in Appendix \@ref{model-dev}. The questions that we ask of the model are as follows:

@@

Before we dive into the answering questions, we provide some background on model interpretation techniques. As the literature is quite sizeable and fast moving, we do not attempt a full taxomony of techniques. The goal is to orient ourselves among broad categories of techniques, so we can map them to tasks indicated by the questions being asked. Model interpretation techniques can be categorized across two dimensions: intrinsic vs. post-hoc and global vs. local.

@@

Intrinsic model interpretation draws conclusions from the structure of the fitted model and are what we typically associate with "interpretable" classes of models. This is only viable with models with simple structures, such as the sparse linear model and shallow decision tree we see in Section \ref{linear-models}, where we arrive at explanations by reading off parameter estimates or a few decision rules. For algorithms that produce models with complex structure that do not lend themselves easily to intrinsic exploration, we can appeal to post-hoc techniques. This class of techniques interrogate the model by presenting it with data for and observing the prediction behavior of the model. These techniques are concerned with only the inputs and outputs, and hence are *agnostic* of the model itself, which means they can also be applied to simple models. Since most useful ML models have a level of complexity beyond the threshold of intrinsic interpretability, we focus on model-agnostic techniques in our case study. As we will see later on, the data that we present to the models are usually some perturbed variations of test data.

@@

Along the other dimension, we categorize model interpretations as global, or model-level, and local, or instance-level. The former class provides insights with respect to the model as a whole. Some examples of these eplanations include variable importances and sensitivities, on average, of the predicted response with respect to individual predictors. In our case study, questions 1 and 2 are associated with this category. On the other hand, question 3 pertains to an individual prediction, which would fall in the local interpretation category. In addition to individual variable attribution, we can also inquire about what would happen to the current predicted response if we were to perturb specific predictor variables.

@@

We follow the methodology of permutation feature importance as described in @fisherAllModels2018, and utilize the notation introduced by @biecekPMVEE. The gist of the technique is as follows: to see how important a variable is, we make predictions without it and see how much worse off we are in terms of accuracy. One way to achieve this would be to re-fit the model many times (as many times as the number of variables.) However, this may be intractable with lengthy model training times or large number of variables, so a more popular approach is to instead keep the same fitted model but permute the values of each predictor.

More formally, let $y$ denote the vector of responses, $X$ denote the matrix of predictor variables, $\widehat{f}$ denote the fitted model, and $L = \mathcal{L}(\widehat{f}(X), y)$ denote the value of the loss function, which is mean squared error in the case of regression. Now, if $\widetilde{X}^{j}$ denotes the predictor matrix where the $j$th variable has been permutated, then we can compute the loss with the permuted dataset as $L^{-j} = \mathcal{L}(\widehat{f}(\widetilde{X}^{j}), y)$. With this, we define the absolute variable importance $VI^{j}$ as $L^{-j} - L$ or the relative variable importance as $VI_{R}^{j} = \frac{L^{-j}}{L}$.

@@

where $N$ is the number of records in our dataset.

In Figure \@ref(fig:pdp), we exhibit the PDP for the "vehicle age" variable. We see that the average predicted loss cost decreases with vehicle age until the latter is around 18. Note that the ticks on the $x$-axis tell us that the data is quite thin for vehicle age greater than 18, so the apparent upward trend to the right is driven by just a few data points. With this in mind, we can conclude that as cars get older, we expect the loss cost predictions generated by our model to decrease.

@@

In the previous two examples, we look at model-level explanations; now we move on to one where we investigate one particular prediction instance. As before, we consider how we would approach the question for linear models. For a GLM with a log link common in ratemaking applications, for example, we simply start with the base rate, then the exponentiated coefficients would have multiplicative effects on the final rate. Similar to the previous examples, for ML models in general we do not have directly interpretable weights. Instead, one way to arrive at variable contributions is calculating the change in the expected model prediction for each predictor, conditioned on other predictors.

@@

\begin{equation}
\widehat{f}(x_* ) = v_0 + \sum_{j=1}^p v(j, x_*),
\end{equation}

where $v(j, x_*)$ denotes the contribution of the $j$th variable in instance $x_*$, defined as

\begin{align}\label{eq:var-contrib}
& v(j, x_* ) = E_X[\widehat{f}(X) | X^1 = x_* ^1, \dots, X^j = x_* ^j] \nonumber\\
& \quad - E_X[\widehat{f}(X^1 = x_* ^1, \dots, X^{j-1} = x_*^{j - 1})].
\end{align}

@@

In Figure \@ref(fig:breakdown) we exhibit a waterfall plot of variable contributions. The "intercept" value denotes the average model prediction. The predicted loss cost for this particular policyholder is slightly less than average; the characteristics that makes this policyholder more risky are the fact that he is a male between the ages of 18 and 25; counteracting the risky driver characteristics are the vehicle properties: it is a GM vehicle built domestically and is seven years old.

>>>>

Proposed by @hiltonConversationalProcesses1990 and interpreted by @millerExplanationArtificial2017 in the context of ML, model explanations are conversations or social interactions. One consequence of this identification is that explanations need to be *relevant* to the *audience*. In developing, filing, and operationalizing a pricing model, one needs to accommodate a variety of stakeholders, each of whom has a different set of questions, assumptions, and technical capacity. First, there are internal stakeholders at the company, which includes management and underwriters. While some of the individuals in this audience may be technical, they are likely less familiar with predictive modeling techniques than the actuaries and data scientists who build the models. Next, we have the regulators, who may have limited resources to review the models, and will focus on a specific list of questions motivated by statute and public policy. Finally, we have the public, who has an interest (perhaps more so than the other parties) because their lives are directly affected by the premiums charged.

It is interesting to note that, the modelers, who are most familiar with the models, tend to be same people designing and communicating the explanations. This poses a challenge that @millerExplainableAI2017 calls "inmates running the asylum", where the modelers design explanations for *themselves* rather than the intended audience. For example, they may be interested in technical questions such as extrapolation behavior and shape the explanations accordingly, which may be lost on the prospective policyholder.

@@

Now that we have established a framework for model interpretation in the form of asking and answering relevant questions, we demonstrate examples of such exchanges via an illustrative case study. Analytically, our starting point is a fitted deep neural network model for predicting loss costs. As the modeling details are of secondary importance, they are available in Appendix \ref{model-dev}. The questions that we ask of the model are as follows:

@@

Before we dive into the answering questions, we present a brief taxonomy of ML interpretation techniques. Rather than attempting an exhaustive classification, the goal is to orient ourselves among broad categories of techniques, so we can map them to tasks indicated by the questions being asked. For our purposes, model interpretation techniques can be categorized across two dimensions: intrinsic vs. post-hoc and global vs. local.

@@

Intrinsic model interpretation draws conclusions from the structure of the fitted model and are what we typically associate with "interpretable" classes of models. This is only viable with models with simple structures, such as the sparse linear model and shallow decision tree we see in Section \ref{linear-models}, where we arrive at explanations by reading off parameter estimates or a few decision rules. For algorithms that produce models with complex structure that do not lend themselves easily to intrinsic exploration, we can appeal to post-hoc techniques. This class of techniques interrogate the model by presenting it with data for scoring and observing the prediction behavior of the model. These techniques are concerned with only the inputs and outputs, and hence are *agnostic* of the model itself, which means they can also be applied to simple models. Since most useful ML models have a level of complexity beyond the threshold of intrinsic interpretability, we focus on model-agnostic techniques in our case study. As we will see later on, the data that we present to the models are usually some perturbed variations of test data.

@@

Along the other dimension, we categorize model interpretations as global, or model-level, and local, or instance-level. The former class provides insights with respect to the model as a whole. Some examples of these eplanations include variable importances and sensitivities, on average, of the predicted response with respect to individual predictors. In our case study, questions 1 and 2 are associated with this category. On the other hand, question 3 pertains to an individual prediction, which would fall in the local, or instance-level, category. In addition to individual variable attribution, we can also inquire about what would happen to the current predicted response if we were to perturb specific predictor variables.

@@

We follow the methodology of permutation feature importance as described in @fisherAllModels2018, and utilize the notation introduced by @biecekPMVEE. The gist of the technique is as follows: to see how important a variable is, we make predictions without it and see how much worse off we are in terms of accuracy. One way to achieve this would be to re-fit the model many times (as many times as the number of variables.) However, this may be intractable with lengthy model training times or large numbers of variables, so a more popular approach is to instead keep the same fitted model but permute the values of each predictor.

More formally, let $y$ denote the vector of responses, $X$ denote the matrix of predictor variables, $\widehat{f}$ denote the fitted model, and $L = \mathcal{L}(\widehat{f}(X), y)$, where $\widehat{f}$ applies to $X$ rowwise, denote the value of the loss function, which is mean squared error in the case of regression. Now, if $\widetilde{X}^{j}$ denotes the predictor matrix where the $j$th variable has been permutated, then we can compute the loss with the permuted dataset as $L^{-j} = \mathcal{L}(\widehat{f}(\widetilde{X}^{j}), y)$. With this, we define the  variable importance $VI^{j}$ as $L^{-j} - L$.

@@

where $N$ is the number of records in the dataset.

In Figure \@ref(fig:pdp), we exhibit the PDP for the "vehicle age" variable. We see that the average predicted loss cost decreases with vehicle age until the latter is around 18. Note that the rug on the $x$-axis tell us that the data is quite thin for vehicle age greater than 18, so the apparent upward trend to the right is driven by just a few data points. With this in mind, we can conclude that as cars get older, we expect the loss cost predictions generated by our model to decrease.

@@

In the previous two examples, we look at model-level explanations; now we move on to one where we investigate one particular prediction instance. As before, we consider how we would approach the question for linear models. For a GLM with a log link common in ratemaking applications, for example, we start with the base rate, then the exponentiated coefficients would have multiplicative effects on the final rate. Similar to the previous examples, for ML models in general we do not have directly interpretable weights. Instead, one way to arrive at variable contributions is calculating the change in the expected model prediction for each predictor, conditioned on other predictors.

@@

\begin{equation}\label{eq:contrib-decompose}
\widehat{f}(x_* ) = v_0 + \sum_{j=1}^p v(j, x_*),
\end{equation}

where $v_0$ denotes the average model response, and $v(j, x_*)$ denotes the contribution of the $j$th variable in instance $x_*$, defined as

\begin{align}\label{eq:var-contrib}
& v(j, x_* ) = E_X[\widehat{f}(X) | X^1 = x_* ^1, \dots, X^j = x_* ^j] \nonumber\\
& \quad - E_X[\widehat{f}(X) | X^1 = x_* ^1, \dots, X^{j-1} = x_*^{j - 1})].
\end{align}

@@

In Figure \@ref(fig:breakdown), we exhibit a waterfall plot of variable contributions. The "intercept" value denotes the average model prediction and represents the $v_0$ term in Equation \@ref(eq:contrib-decompose). The predicted loss cost for this particular policyholder is slightly less than average; the characteristics that makes this policyholder more risky are the fact that he is a male between the ages of 18 and 25; counteracting the risky driver characteristics are the vehicle properties: it is a GM vehicle built domestically and is seven years old.
